Spatial Reasoning and Data Displays

### 2015-01-27 Paper/Draft1/TestingVisualAptitude.Rnw
Influence plot explanation

<<lineupblockpca, echo=F, warning=F, message=F, fig.cap="Lineup block PCA rotation matrix in graphical form.", fig.width=5, fig.height=5>>=
lineup.block.pca <- prcomp(lineup.section.summary[,c("lineup_1", "lineup_2", "lineup_3", "card_rot", "fig_class", "folding", "vis_search")], scale=T, retx=T)
rotation <- melt(lineup.block.pca$rotation)
importance <- as.data.frame(summary(lineup.block.pca)$importance)[2,]
rotation <- merge(rotation, melt(importance, variable.name="Var2", value.name="importance"))
rotation$Var1 <- factor(rotation$Var1, 
                        levels=c("vis_search", "folding", "card_rot", "fig_class", "lineup_3", "lineup_2", "lineup_1"),
                        labels=c("Visual Search", "Paper Folding", "Card Rotation", "Figure Class.", "Lineup Test 3", "Lineup Test 2", "Lineup Test 1"))
library(grid) # needed for arrow function
qplot(x=0,  y=Var1, xend=value*importance, yend=Var1, geom="segment", data=rotation, arrow=arrow(length=unit(0.1, "cm"))) + facet_wrap(~Var2, ncol=3) + xlab("Contribution to each PC * % Variance in PC") + ggtitle("Lineup Types and Visual Aptitude") + ylab("") + theme_bw()
@

% \comment{add something about x and y axis in the description of the figure. }
\comment{Let's give this plots a name ... how about {\it PC importance plots}? Get rid of the * in the x axis title. Both `realtive contribution' and `importance' are a bit fuzzy terms ... I'm not sure what you are doing until I'm looking at the code. Define importance and each of the items, and yes, that might need a bit of math.}
\newtext{Figure~\ref{fig:lineupblockpca} shows the relative contribution of each input variable  to each principal component (PC), weighted by the importance of each principal component. The input variables are shown on the $y$ axis, with the importance to each PC shown on the $x$ axis. This allows us to consider the rotation matrix visually; for instance, we see that again, the first PC accounts for most of the variance and seems to represent general visual aptitude. }
% PC2 emphasizes the relationship between figure classification and lineup task 1 (and to a lesser extent, on lineup task 2). PC3 differentiates the lineup task 3 from the visual search task; other tasks do not impact PC3 to a meaningful degree. PC4 separates lineup test 2 from the paper folding and card rotation tests and emphasizes the overlap with the visual search test. PC5 separates lineup test 2 and the figure classification test from the overlap between the other lineup tests and the visual search test. PC6 and PC7 account for a joint total of less than 10\% of the overall variability, and do not display any distinct patterns in the loadings.
%
PC2 emphasizes the overlapping variation in performance on lineup test 1 and the figure classification test. PC3 emphasizes the additional variation in performance on lineup test 3 and the visual search test, while PC4 emphasizes the extra variability in performance on lineup test 2 and the paper folding test. All three lineups, plus the figure classification, paper folding, and visual search tests contribute to PC5. PC6 and PC7 jointly account for less than 10\% of the variance in the data and do not display any distinct patterns in the loadings. 
%\Sexpr{as.numeric(round(100*importance[6], 1))}\% and \Sexpr{as.numeric(round(100*importance[7], 1))}\% of the variance, respectively; PC6 seems to differentiate the first two lineup tasks from the overlap between the third lineup task and the figure classification task.

>>>>

\newtext{Principal component analysis separates multivariate data into orthogonal components using a rotation matrix, which contains information on the contribution of each input variable to the principal component. The importance of a principal component is also evaluated to assess the proportion of the overall variance in the data contained within the component (table \ref{tab:PCAvariance5} shows the importance of each PC for the analysis of the aggregate lineup score and the four aptitude tests).}

<<lineupblockpca, echo=F, warning=F, message=F, fig.cap="Principal Component importance plot, showing each input variable's contribution to the principal component, scaled by the proportion of variability in the data contained in the principal component.", fig.width=5, fig.height=5>>=
lineup.block.pca <- prcomp(lineup.section.summary[,c("lineup_1", "lineup_2", "lineup_3", "card_rot", "fig_class", "folding", "vis_search")], scale=T, retx=T)
rotation <- melt(lineup.block.pca$rotation)
importance <- as.data.frame(summary(lineup.block.pca)$importance)[2,]
rotation <- merge(rotation, melt(importance, variable.name="Var2", value.name="importance"))
rotation$Var1 <- factor(rotation$Var1, 
                        levels=c("vis_search", "folding", "card_rot", "fig_class", "lineup_3", "lineup_2", "lineup_1"),
                        labels=c("Visual Search", "Paper Folding", "Card Rotation", "Figure Class.", "Lineup Test 3", "Lineup Test 2", "Lineup Test 1"))
library(grid) # needed for arrow function
qplot(x=0,  y=Var1, xend=value*importance, yend=Var1, geom="segment", data=rotation, arrow=arrow(length=unit(0.1, "cm"))) + facet_wrap(~Var2, ncol=3) + xlab("Influence") + ggtitle("Lineup Types and Visual Aptitude") + ylab("") + theme_bw()
@

% \comment{add something about x and y axis in the description of the figure. }
% \comment{Let's give this plots a name ... how about {\it PC importance plots}? Get rid of the * in the x axis title. Both `realtive contribution' and `importance' are a bit fuzzy terms ... I'm not sure what you are doing until I'm looking at the code. Define importance and each of the items, and yes, that might need a bit of math.}
Figure~\ref{fig:lineupblockpca} shows the influence of each input variable on each principal component; the arrows are scaled in proportion to the importance of each principal component. The input variables are shown on the $y$ axis, with the influence of the variable on each PC shown on the $x$ axis. This allows us to consider the rotation matrix visually; for instance, we see that again, the first PC accounts for most of the variance and seems to represent general visual aptitude. 
% PC2 emphasizes the relationship between figure classification and lineup task 1 (and to a lesser extent, on lineup task 2). PC3 differentiates the lineup task 3 from the visual search task; other tasks do not impact PC3 to a meaningful degree. PC4 separates lineup test 2 from the paper folding and card rotation tests and emphasizes the overlap with the visual search test. PC5 separates lineup test 2 and the figure classification test from the overlap between the other lineup tests and the visual search test. PC6 and PC7 account for a joint total of less than 10\% of the overall variability, and do not display any distinct patterns in the loadings.
%
PC2 emphasizes the overlapping variation in performance on lineup test 1 and the figure classification test. PC3 emphasizes the additional variation in performance on lineup test 3 and the visual search test, while PC4 emphasizes the extra variability in performance on lineup test 2 and the paper folding test. All three lineups, plus the figure classification, paper folding, and visual search tests contribute to PC5. PC6 and PC7 jointly account for less than 10\% of the variance in the data and do not display any distinct patterns in the loadings. 
%\Sexpr{as.numeric(round(100*importance[6], 1))}\% and \Sexpr{as.numeric(round(100*importance[7], 1))}\% of the variance, respectively; PC6 seems to differentiate the first two lineup tasks from the overlap between the third lineup task and the figure classification task.
